---
title: Tidy Tuesday - Plants in Danger
author: Richard Bamattre
date: "2023-01-01"
output: html_notebook
---

**Purpose**



**Description**



**Data Sources**



*Other Notes*



*Change log*


```{r setup}
## Load libraries

# read
library(tidytuesdayR)

# explore

# wrangle
library(tidyverse) # key packages - remove if not using
library(lubridate)
library(janitor)

# model

# viz
library(biscale)
library(cowplot)
library(hrbrthemes)

#extrafont::font_import() # install new fonts
extrafont::loadfonts(device = "win")

# export

## Custom functions

# If not using a project, set the working directory to where the markdown
# script is located:

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Read

```{r read}
tuesdata <- tt_load("2020-08-18")

plants <- tuesdata$plants

threats <- tuesdata$threats

actions <- tuesdata$actions
```

# Initial cleaning

The plants dataframe is in wide/untidy format, while threats and actions are long/tidy. Threats and action are sparse datasets, meaning threats and actions that are not relevant to a plant are included but coded as zero.

Let's make these less sparse by remove those zeroes.

```{r}
threats2 <- threats |> 
  filter(threatened == 1) |> 
  select(-threatened)

actions2 <- actions |> 
  filter(action_taken == 1) |> 
  select(-action_taken)
```

# Explore

Let's look at the data using crosstables to see what variables and categories look like, what may be related.

```{r explore}
tabyl(threats2, continent)
```

```{r}
tabyl(threats2, group)
```

```{r}
tabyl(threats2, red_list_category)
```

```{r}
tabyl(threats2, threat_type) |> 
  arrange(desc(n))
```

```{r}
threats2 |> 
  count(country, threat_type) |> 
  arrange(desc(n))
```


## Get world data

May be interesting to see a map, for exploration or the final plot.

Load in world data

```{r}
bkg_color <- "#f5ebe0"

# summarize per country

threats_country <- threats2 |> 
  group_by(country) |> 
  summarize(threats = n_distinct(binomial_name))

# new for me - ggplot function to get a map of the world

world <- map_data("world") |> 
  filter(region != "Antarctica")

# use geom_map because we can map and visualize data without having to combine

g <- ggplot() +
  geom_map(data = world, map = world, aes(long, lat, map_id = region),
           fill = bkg_color) + # base map
  geom_map(data = threats_country, map = world,
           aes(fill = threats, map_id = country)) + # color filled map on top
  coord_quickmap() + # quickly use a standard projection
  scale_fill_viridis_c() +
  labs(title = "Threats to plants per country") +
  theme_void() +
  theme(text = element_text(family = "IBM Plex Sans"))

g
```

## Correlations

```{r}
corrplot::corrplot(cor(plants[,6:23]))
```

It would be interesting to visualize the types of threats impacting plants. But currently it's a pretty high dimensionality (lots of different threats). Would a PCA be useful here?

```{r}
pca <- prcomp(plants[,6:16]) # only threats and remove NA

summary(pca)
```

```{r}
factoextra::fviz_eig(pca)
```

```{r}
factoextra::fviz_pca_var(pca)
```

Natural Systems Modifications + Biological Resources Use; Commercial Development + Agriculture and Aquaculture

Apparently we can't specify the number of principal components? So we can do a factor analysis

```{r}
fact <- factanal(plants[,6:16], 2, scores = "Bartlett")

summary(fact)
```

```{r}
fact$loadings
```

```{r}
plants2 <- plants |> 
  cbind(fact$scores)
```

Factor 1: Climate change, geological events (less agriculture, biological resource use)
Factor 2: Commercial development, pollution, human intrusions, transportation corridor

```{r}
ggplot(data = plants2, aes(x = Factor1, y = Factor2, color = continent)) +
  geom_point()
```


## Bivariate map

Now we have two overlapping factors to visualize. Can we use bivarate mapping to do this?

https://cran.r-project.org/web/packages/biscale/vignettes/biscale.html

Scores are at plant level, any better way to summarize this besides a mean?

```{r}
country_data <- plants2 |> 
  group_by(country) |> 
  summarize(Factor1 = mean(Factor1),
            Factor2 = mean(Factor2))
```


```{r}
bi_data <- bi_class(country_data, 
                    x = Factor1, y = Factor2, style = "quantile", dim = 3)
```

```{r}
b <- ggplot() +
  geom_map(data = world, map = world, aes(long, lat, map_id = region),
           fill = "white") + # base map
  geom_map(data = bi_data, map = world,
           aes(fill = bi_class, map_id = country),
           show.legend = FALSE) + # color filled map on top
  coord_quickmap() + # quickly use a standard projection
  bi_scale_fill(pal = "DkViolet2", dim = 3) +
  labs(title = "Threats to plant species") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, color = "#9b2226",
                                  family = "IBM Plex Sans SemiBold"),
        text = element_text(family = "IBM Plex Sans"),
        plot.background = element_rect(bkg_color, color = NA))

legend <- bi_legend(pal = "DkViolet2",
                    dim = 3,
                    xlab = "Climate change threats",
                    ylab = "Industrial threats",
                    size = 6) +
  theme(text = element_text(family = "IBM Plex Sans"),
        plot.background = element_rect(fill = bkg_color, color = NA),
        panel.background = element_rect(fill = bkg_color, color = NA))

# cowplot allows us to draw multiple plots on a single canvas

final <- ggdraw() +
  draw_plot(b, 0, 0, 1, 1) +
  draw_plot(legend, 0, .1, 0.3, 0.3) +
  draw_text("Tidy Tuesday | @rbamattre | Source: IUCN Red List of Threatened Spcies (2020)", size = 8, x = .98, y = 0.05, hjust = 1, family = "IBM Plex Sans")

final

ggsave(plot = last_plot(), device = "png", dpi = 200, bg = bkg_color,
       width = 8, height = 5,
       filename = "2020-08-18_plants_in_danger_final_rb.png")
```

# Wrangle

```{r wrangle}

```

# Model

```{r model}

```

# Export

```{r export}

```