---
title: Tidy Tuesday - US Droughts
author: Richard Bamattre
date: "2023-12-29"
output: html_notebook
---

**Purpose**

**Description**

**Data Sources**

*Other Notes*

*Change log*

```{r setup}
## Load libraries

# read
library(tidytuesdayR)

# explore

# wrangle
library(tidyverse) # key packages - remove if not using
library(lubridate)

# viz
library(extrafont)

# model

# export

## Custom functions

loadfonts()

# If not using a project, set the working directory to where the markdown
# script is located:

#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Read

```{r read}
dr <- tt_load('2021-07-20')$drought
```

# Explore

```{r explore}
min(dr$valid_start); max(dr$valid_start)
```

```{r}
dr |> 
  group_by(state_abb) |> 
  summarize(rows = n())
```
```{r}
dr |> 
  filter(stat_fmt == 1) # we only have categorical data?
```


```{r}
dr |> 
  filter(state_abb == "CA",
         valid_start >= "2020-01-01") |> 
  ggplot(aes(x = valid_start, y = area_pct, group = state_abb, color = drought_lvl)) +
    geom_line() +
    facet_wrap(~ drought_lvl) +
    guides(color = "none")
```

For a given year, each state would have weekly? values for None, Drought Levels 0-4

```{r}
# define a drought color palette

colfunc <- colorRampPalette(c("#ED9654", "#1E2E3C"))

colfunc(5)

pal <- c(colfunc(5), "gray") # include color for "None

# using color picker for data: http://tristen.ca/hcl-picker/#/hlc/6/1.05/1E2E3C/ED9654

pal2 <- c("#1E2E3C", "#0D524F", "#3D7346", "#8F8C36", "#ED9654", "gray")

# including None in the pallette

pal3 <- c("#0E4B4E", "#23674D", "#5B7F3E", "#A18F37", "#ED9654", "#1E2E3C")
```


```{r, fig.width = 10}
g <- dr |> 
  filter(state_abb == "CA") |> 
  ggplot(aes(x = valid_start, y = area_pct, group = state_abb, fill = drought_lvl)) +
    geom_col(width = 7) # set width to 7 days

g + scale_fill_manual(values = pal)
```

```{r, fig.width = 10}
g +
  scale_fill_manual(values = pal2)
```

```{r, fig.width = 10}
g +
  scale_fill_manual(values = pal3)
```

For each time point, there is a number for each drought level (d0 -> d4 more extreme drought). So each state has 6 rows for each time point (weekly periods)

```{r}
dr |> 
  filter(drought_lvl == "D4",
         valid_start >= "2020-01-01") |> 
  ggplot(aes(x = valid_start, y = area_pct, group = state_abb, color = state_abb)) +
    geom_line() +
    facet_wrap(~ drought_lvl) +
    guides(color = "none")
```


# Wrangle

```{r wrangle}

```

# Viz

Resource for alternatives to black (to highlight a sequential color scheme like viridis or the one I'm using here): https://squarebase.co/blogs/news/alternative-colours-to-black-to-use-on-your-squarespace-website

```{r model, fig.width = 10}
library(geofacet)

geo_plot <- function(data, year) {

  data |> 
    filter(year(valid_start) == year) |> 
  ggplot(aes(x = valid_start, y = area_pct, group = state_abb, fill = drought_lvl)) +
    geom_col(width = 7) +
    scale_fill_manual(values = pal3) +
    #scale_x_date(expand = c(0.1, 0.1)) +
    facet_geo(~ state_abb, grid = "us_state_grid1") +
    theme_minimal() +
    theme(text = element_text(color = "bisque1",
                              family = "Roboto"),
          strip.text = element_text(color = "bisque4",
                                    family = "Roboto"),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          #panel.background = element_rect(fill = "darkgray"),
          plot.background = element_rect(fill = "#1f1206")) +
    labs(fill = "Drought Level", x = "", y = "",
         title = paste("Drought Levels for", year),
         caption = "Tidy Tuesday | @rbamattre | Source: U.S. Drought Monitor")
  
}

geo_plot(dr, 2020)

#geo_plot(dr, 2001)
```

# Animate

```{r}
# get a vector of all years in the data
years <- seq(2001, 2021, by = 1)
```


```{r animate}
library(animation)

# use animation and purrr to iterate function for each year and save output as gif

animation::saveGIF(
  expr = {
    purrr::walk(
      years,
      ~ plot(geo_plot(dr, .))
    )
  },
  movie.name = "us_droughts_rb.gif",
  ani.width = 800,
  ani.height = 500
)
```
