---
title: Tidy Tuesday LEGO
author: Richard Bamattre
date: "2023-01-01"
output: html_notebook
---

**Purpose**



**Description**



**Data Sources**



*Other Notes*



*Change log*


```{r setup}
## Load libraries

# read
library(tidytuesdayR)

# wrangle
library(tidyverse) # key packages - remove if not using
library(lubridate)
library(janitor)

# viz
library(ggthemes)
library(magick)
library(extrafont)

# model

# export

## Custom functions

loadfonts()
```

# Read

From rebrickable

The schema from here is useful: https://rebrickable.com/downloads/

```{r}
#tuesdata <- tt_load('2022-09-06')

# getting some issue with vroom

inventories <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/inventories.csv.gz')

inventory_sets <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/inventory_sets.csv.gz')

sets <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/sets.csv.gz')

themes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/themes.csv.gz')

inv_parts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/inventory_parts.csv.gz')

colors <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-06/colors.csv.gz') %>%
  mutate(hex = paste0("#", rgb))

```

# Explore

I think inventories are used to connect the sets with the parts. I'm not sure I need to get that granular at the moment.

```{r}
summary(sets)
```

```{r}
sets |> 
  filter(num_parts == 0)
```

Games and other products are 0 part sets

```{r}
sets2 <- sets |> 
  filter(num_parts > 0) |> 
  left_join(themes |> rename(theme = name), join_by(theme_id == id)) %>%
  left_join(themes |> rename(parent_theme = name) |> 
              select(-parent_id),
            join_by(parent_id == id)) |> 
  mutate(overall_theme = coalesce(parent_theme, theme))

sets2
```

```{r}
tabyl(sets2, overall_theme) |> 
  arrange(desc(n)) |> 
  adorn_pct_formatting()
```

```{r}
sets2 |> 
  ggplot(aes(x = year, y = num_parts)) +
    geom_jitter(alpha = 0.4, stroke = NA) +
    geom_smooth()
```

```{r}
sets2 |> 
  arrange(desc(num_parts)) |> 
  head(n = 100)
```

Get the most common themes (top 25?)

```{r}
top_25 <- sets2 |> 
  group_by(overall_theme) |> 
  summarize(sets = n_distinct(set_num)) |> 
  arrange(desc(sets)) |> 
  slice(1:10)

top_25
```

```{r}
sets2 |> 
  filter(overall_theme %in% top_25$overall_theme) |> 
  group_by(overall_theme, year) |> 
  summarize(sets = n_distinct(set_num),
            avg_parts = mean(num_parts)) |> 
  ggplot(aes(x = year, y = sets, fill = overall_theme)) +
    geom_col(
      #position = "fill"
      ) +
    scale_fill_ptol()
```

```{r}
sets2 |> 
  filter(overall_theme %in% top_25$overall_theme) |> 
  group_by(overall_theme, year) |> 
  summarize(sets = n_distinct(set_num),
            avg_parts = mean(num_parts)) |> 
  ggplot(aes(x = year, y = avg_parts, fill = overall_theme)) +
    geom_col(
      #position = "fill"
      ) +
    scale_fill_ptol() +
    facet_wrap(~ overall_theme) +
    theme(legend.position = "none")
```

```{r}
sets2 %>%
  filter(overall_theme == "Service Packs")
```

```{r}
sets2 %>%
  filter(overall_theme == "Educational and Dacta")

image_read(sets2[51,]$img_url) |> print() # print the picture for that row



image_read(sets2$img_url[sets2$set_num == "970608-1"]) |> print()
```


# Wrangle

Add parts + Colors

```{r wrangle}
sets_parts <- sets2 %>%
  left_join(inventories, join_by(set_num),
            multiple = "all") %>%
  left_join(inv_parts %>%
              rename(part_img_url = img_url), 
            join_by(id == inventory_id),
            multiple = "all") %>%
  left_join(colors %>%
              rename(color_name = name), join_by(color_id == id))
```

```{r}
# get picture for a set (even if there's multiple rows per parts)

#image_read(sets_parts[sets_parts$set_num == '31120-1',][1,]$img_url) |> print()
# this is odd base R subsetting
```


# Color graph function

Create a function that, when inputted any set number, will map the colors used

```{r}
library(treemapify)

plot.set <- function(sets, type = "bar", show = "name") {
  
  # filter the data to that set
  
  set_data <- sets_parts %>%
    filter(set_num %in% sets)
  
  if (show == "theme") {
    
    set_data <- set_data %>%
      mutate(name = overall_theme)
    
  }
  
  # summarize the number of parts per color in that set
  
  set_colors <- set_data %>%
    group_by(set_num, name, year, color_id) %>%
    summarize(parts = n_distinct(part_num),
              hex = min(hex), .groups = "drop") %>%
    arrange(desc(parts)) %>%
    mutate(row = row_number())
  
  # create a named vector of color values
  
  color_vals <- as.character(set_colors$hex)
  names(color_vals) <- as.character(set_colors$color_id)
  
  # plot!
  
  if (type == "bar") {
  
    set_colors |> 
      ggplot(aes(x = 1, y = parts, fill = as.factor(color_id))) +
        geom_bar(stat = "identity") +
        coord_flip() +
        scale_fill_manual(values = color_vals) +
        theme_minimal() + 
        theme(legend.position = "none",
              axis.text = element_blank(),
              axis.title = element_blank(),
              panel.grid = element_blank(),
              panel.background = element_rect(fill = "gray", color = NA)) +
        facet_wrap(~ name, scales = "free")
    
  }
  
  else if (type == "tree") {
    
    set_colors |> 
      ggplot(aes(area = parts, fill = as.factor(color_id))) +
        geom_treemap() +
        scale_fill_manual(values = color_vals) +
        theme_minimal() + 
        theme(legend.position = "none",
              axis.text = element_blank(),
              axis.title = element_blank(),
              panel.grid = element_blank()) +
        facet_wrap(~ name, scales = "free")
    
  }
  
  else if (type == "pie") {
    
    set_colors |> 
      ggplot(aes(x = "", y = parts, fill = as.factor(color_id))) +
        geom_col(color = "lightgray") +
        coord_polar(theta = "y") +
        scale_fill_manual(values = color_vals) +
        theme_minimal() + 
        theme(legend.position = "none",
              axis.text = element_blank(),
              axis.title = element_blank(),
              panel.grid = element_blank()) +
        facet_wrap(~ name)
    
  }

}
```

```{r}
top_sets <- sets2 %>%
  arrange(desc(num_parts)) %>%
  slice(1:10) %>%
  pull(set_num)

plot.set(top_sets, type = "bar")

plot.set(top_sets, type = "tree")

plot.set(top_sets, type = "pie") # scaling doesn't work but pies are not great anyways
```

Get an example from popular themes - set with most parts

```{r}
themes <- sets2 %>%
  group_by(overall_theme) %>%
  mutate(sets = n_distinct(set_num)) %>%
  filter(num_parts == max(num_parts)) %>% # get the biggest set per theme
  ungroup() %>%
  arrange(desc(sets)) %>%
  slice(1:30)
```

```{r}
p <- plot.set(themes$set_num, type = "tree", show = "theme") +
  labs(title = "LEGO colors for set with the most parts, by category",
       caption = "Tidy Tuesday | @rbamattre | Source: rebrickable") +
  theme(text = element_text(family = "Bahnschrift"), legend.position = "none")

p

ggsave("2022-09-06 lego.png", p, device = "png", dpi = 150,
       width = 8, height = 6, bg = "white")
```



# Model

```{r model}

```

# Export

```{r export}

```