---
title: Tidy Tuesday - Animal Rescues
author: Richard Bamattre
date: "2021-12-07"
output: html_notebook
---

**Purpose**

https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-06-29/readme.md


```{r setup}
## Load libraries

# read
library(tidytuesdayR)

# wrangle
library(tidyverse)
library(lubridate)
library(janitor)
library(tidytext)

# viz
library(ggthemes) # ggplot themes
library(ggtext) # html/markdown styling ggplot text
library(extrafont)
library(gridExtra)
library(grid)

# model
library(sf) # simple features for GIS and mapping

# Necessary if there are issues of no font name loading in extrafont
# https://stackoverflow.com/questions/61204259/how-can-i-resolve-the-no-font-name-issue-when-importing-fonts-into-r-using-ext
#remotes::install_version("Rttf2pt1", version = "1.3.8")

# export

## Custom functions

#font_import(paths = "C:/Windows/Fonts") # load fonts for the first time, takes a few minutes

extrafont::loadfonts(device = "win") # 

# set working directory to wherever the RMD is located
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Read

```{r read}
tuesdata <- tidytuesdayR::tt_load('2021-06-29')

ani <- tuesdata$animal_rescues
```

# Wrangle

```{r wrangle}
ani2 <- ani %>%
  mutate(across(is.character, ~ na_if(.x, "NULL"))) %>% # turn NULLs into NA
  mutate(date_time_of_call = mdy_hm(date_time_of_call)) %>% # make into datetime
  mutate(across(latitude:longitude, as.numeric)) # convert coordinates to numeric
```

# Explore

What kind of info is in animal group?

```{r}
table(ani2$animal_group_parent)

# cat in two categories. What's a budgie?

ani3 <- ani2 %>%
  mutate(animal_group_parent = factor(animal_group_parent) %>%
           recode(cat = "Cat", # change values
                  Budgie = "Bird",
                  Pigeon = "Bird",
                  Lamb = "Sheep",
                  Bull = "Cow"))

table(ani3$animal_group_parent)
```

How many records have coordinates?

```{r}
sum(!is.na(ani3$latitude)) / nrow(ani3)
```
Can we use the UK postcode instead?

```{r}
sum(!is.na(ani3$postcode_district)) / nrow(ani3)
```

Import dataset of UK postcodes, get district postcodes

```{r}
uk_pc <- read_csv("ONSPD_MAY_2016_UK.csv")

# just get postcode district, not sector or unit
# https://en.wikipedia.org/wiki/Postcodes_in_the_United_Kingdom#Postcode_district

uk_pc2 <- uk_pc %>%
  separate(pcd2, into = c("postcode_district", "pcd_2nd_half")) %>%
    # get district postcode out - however we'll have multiple lat/lon for each
  group_by(postcode_district) %>%
  summarize(lat = mean(lat), # summarize average - not perfect method
            long = mean(long))

ani4 <- ani3 %>%
  left_join(uk_pc2 %>% select(postcode_district, lat, long), by = "postcode_district")

# how accurate are these when we know the specific lat/long?

ani4 %>%
  filter(!is.na(latitude),
         latitude > 0) %>%
  mutate(lat_error = latitude - lat,
         long_error = longitude - long) %>%
  ggplot() +
    geom_density(aes(lat_error), color = "red") +
    geom_density(aes(long_error), color = "blue")

# One degree of lat/lon is about 69 miles, so most aren't far off.
# I think the average lat/lon by postal code district is okay for a rough map
```

# Model

Ideas: make a map of animals by type

```{r model}
ani_geo <- ani4 %>%
  st_as_sf(coords = c("long", "lat"))

plot(ani_geo["ward"])
```

```{r, fig.width=8, fig.height=8}
ani_geo %>%
  filter(animal_group_parent %in% c("Bird", "Cat", "Dog", "Fox", "Deer", "Horse"),
         ) %>%
  ggplot(aes(color = animal_group_parent)) +
    geom_sf() +
    theme(legend.position = "bottom")
```


Points may be hard. Get polygons for London postal codes

```{r}
pc_geo <- read_sf("https://raw.githubusercontent.com/sjwhitworth/london_geojson/master/london_postcodes.json")

plot(pc_geo["Name"])
```

```{r}
ani_geo2 <- ani4 %>%
  left_join(pc_geo, by = c("postcode_district" = "Name")) %>% # join based on pc district
  st_as_sf() # convert to sf object

plot(ani_geo2["ward_code"])
```

Now that we have polygons containing multiple incidents, we could probably summarize.

```{r}
# summarize using non-geo tibble, then add the geometry (otherwise get errors)
# right join here since there are 14 postal districts without data

ani_geo_sum <- ani4 %>%
  count(postcode_district, animal_group_parent) %>%
  right_join(pc_geo, by = c("postcode_district" = "Name")) %>%
  st_as_sf()

ani_geo_sum %>%
  filter(animal_group_parent == "Fox") %>%
  ggplot(aes(fill = n)) +
    geom_sf(color = NA) +
    theme_map() +
    labs(title = "Fox rescues in London") +
    scale_fill_viridis_b()

ani_geo_sum %>%
  filter(animal_group_parent == "Cat") %>%
  ggplot(aes(fill = n)) +
    geom_sf(color = NA) +
    theme_map() +
    labs(title = "Cat rescues in London") +
    scale_fill_viridis_b()
```

Most common animal rescue for each postal district?

```{r}
ani_geo_sum2 <- ani4 %>%
  count(postcode_district, animal_group_parent) %>%
  group_by(postcode_district) %>%
  arrange(postcode_district, desc(n)) %>%
  slice(1) %>%
  droplevels() %>%
    # since animal group parent is still a factor, drop empty values
  right_join(pc_geo, by = c("postcode_district" = "Name")) %>%
  st_as_sf() %>%
  mutate(animal2 = case_when(is.na(n) ~ "No rescues",
                             TRUE ~ as.character(animal_group_parent)))

table(ani_geo_sum2$animal2)

ani_geo_sum2 %>%
  ggplot() +
    geom_sf(aes(fill = animal2), color = NA) +
    theme_map() +
    labs(fill = "Most common animal",
         title = "Animal rescues in London, 2009-2021") +
    scale_fill_brewer(type = "qual", palette = 2) +
    theme(legend.position = "right")
```

Interesting first stab, but lacks nuance. Are all those areas 90% cat or 55%? 

National Geographic's predominant race map does a great job of amounts in categories
https://flowingdata.com/2018/10/12/mapping-predominant-race-block-by-block/

```{r}
ani_geo_sum3 <- ani4 %>%
  count(postcode_district, animal_group_parent) %>%
  group_by(postcode_district) %>%
  mutate(total = sum(n)) %>% # calculate total rescues
  mutate(perc = n / total) %>% # get percent
  filter(!animal_group_parent %in% c("Cat", "Dog")) %>%
  arrange(postcode_district, desc(perc)) %>%
  slice(1) %>%
  droplevels() %>%
    # since animal group parent is still a factor, drop empty values
  right_join(pc_geo, by = c("postcode_district" = "Name")) %>%
  st_as_sf() %>%
  mutate(animal2 = case_when(is.na(n) ~ "No rescues",
                             animal_group_parent == "Unknown - Heavy Livestock Animal" ~
                               "Heavy Livestock Animal",
                             TRUE ~ as.character(animal_group_parent)),
         perc = if_else(animal2 == "No rescues", 1, perc))

table(ani_geo_sum3$animal2)

ani_geo_sum3 %>%
  ggplot(aes(y = animal2, x =  perc, fill = animal2)) +
    geom_boxplot() +
    geom_jitter(alpha = 0.5, aes(color = animal2)) +
    scale_fill_brewer(type = "qual", palette = 2) +
    scale_color_brewer(type = "qual", palette = 2)
```


```{r}
# hue by animal, but brightness percent (dominance)

# set font
font = "Bahnschrift"

# change theme
theme_map2 <- theme_map() +
  theme(text = element_text(family = font))

theme_set(theme_map2)

# get better colors
# calc from ggthemes is nice, but has two greens and blues,
# won't work with shading

# pull out 7 colors

pal <- calc_pal()(6)

# change no rescues color

pal[6] <- "#5b5b5b" #"#ffffff"

scales::show_col(pal)

# change heavy livestock animal color

p <- ani_geo_sum3 %>%
  ggplot() +
    geom_sf(aes(fill = animal2, alpha = rev(perc)), color = NA) +
    #theme_map2() +
    labs(fill = "Most common animal",
         caption = "Color by most common animal, shaded by its prevalence",
         alpha = "Prevalence") +
    scale_fill_manual(values = pal) +
    scale_alpha_continuous(labels = scales::percent_format()) +
    theme(legend.position = "left")

p

# save plot


# bring together the plots

g <- arrangeGrob(p,
                 top = textGrob("No dogs or cats: Other animal rescues in London, 2009-2021", 
                                gp = gpar(fontfamily = font,
                                          fontface = "bold",
                                          col = "black",
                                          hjust = 0,
                                          cex = 1.2)),
                 bottom = textGrob("Tidy Tuesday | @rbamattre | Source: London.gov", 
                                   gp = gpar(fontfamily = font,
                                             col = "black",
                                             hjust = 0,
                                             cex = 0.8)))

ggsave("2020-06-29_animal_rescue_london.png", g, width = 8, height = 6, dpi = 150,
       bg = "white")
```


# Export

```{r export}

```