---
title: Tidy Tuesday - Animal Rescues
author: Richard Bamattre
date: "2022-02-25"
output: html_notebook
---

**Purpose**

https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-01-25/readme.md


```{r setup}
## Load libraries

# read
library(tidytuesdayR)

# wrangle
library(tidyverse) # yup
library(lubridate) # work with dates
library(janitor) # clean up names + nice tables

# viz
library(ggthemes) # ggplot themes
library(ggtext) # html/markdown styling ggplot text
library(extrafont) # use more fonts in ggplot

# model


# Necessary if there are issues of no font name loading in extrafont
# https://stackoverflow.com/questions/61204259/how-can-i-resolve-the-no-font-name-issue-when-importing-fonts-into-r-using-ext
#remotes::install_version("Rttf2pt1", version = "1.3.8")=

# export

## Custom functions

#font_import(paths = "C:/Windows/Fonts") # load fonts for the first time, takes a few minutes

extrafont::loadfonts(device = "win") # 

# set working directory to wherever the RMD is located
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Read

```{r read}
tuesdata <- tidytuesdayR::tt_load('2022-01-25')

rat <- tuesdata$ratings
deet <- tuesdata$details
```

What factors predict most accurately a highly ranked game?

Helpful if ratings are normally distributed.

```{r}
rat %>%
  ggplot(aes(average)) +
    geom_histogram(bins = 100)

rat %>%
  ggplot(aes(bayes_average)) +
    geom_histogram(bins = 100)

# difference between average and bayes average?

rat %>%
  mutate(diff = average - bayes_average) %>%
  ggplot(aes(diff)) +
    geom_histogram(bins = 100)
```

# Wrangle

We need to pull out some of these lists into variables. If we're productionizing a model, we'd use a pipeline. Since it's a simple exercise, let's clean in R (easier for me to quickly) and then use scikit learn in Python to model.

```{r wrangle}
# number of distinct categories/families - this includes combinations though
n_distinct(deet$boardgamecategory)
n_distinct(deet$boardgamefamily) # way too many families to use!

deet2 <- deet %>%
  mutate(category = str_remove_all(boardgamecategory, "\\[|\\]|'")) %>%
  separate_rows(category, sep = ",") %>%
  mutate(category = str_squish(category),
         value = 1) %>%
  select(id, primary, yearpublished:minage, category, value) # select key features

n_distinct(deet2$category) # 85 distinct categories, that's a lot

# Turn into wide data - each category will be a column
# This isn't tidy, but it allows cluster analysis, modeling, etc.

deet3 <- deet2 %>%
  pivot_wider(names_from = "category", values_from = "value") %>%
  mutate(age = 2021 - yearpublished) %>%
  select(-yearpublished) %>%
  select(id, primary, age, everything()) %>%
  mutate(across(Medical:`Fan Expansion`, ~ replace_na(.x, 0))) %>%
  left_join(rat %>% select(id, average), by = "id")
```

# Cluster analysis

Can we create some higher level categories?

```{r}
library(FactoMineR)
library(factoextra)

# get just numeric category variables
cat <- deet3 %>% select(Medical:`Fan Expansion`)

# hierarchical clustering - takes way too long
# hclust <- hclust(dist(cat))

# kmeans
set.seed(123)

km <- kmeans(cat, 12)

# to much data to visualize
#fviz_cluster(km, data = cat)

# visualize centers

km_centers <- km$centers %>% 
  as_tibble %>%
  rownames_to_column("cluster") %>%
  mutate(cluster = as.numeric(cluster)) %>%
  pivot_longer(-cluster, names_to = "category", values_to = "means")

km_centers %>%
  filter(means > .05) %>%
  arrange(cluster, desc(means)) %>%
  ggplot(aes(y = category, x = means, fill = factor(cluster))) +
    geom_col() +
    facet_wrap(~ cluster, scales = "free_y")
  
# PCA - principal component analysis
pca <- PCA(cat, ncp = 40, graph = FALSE)

# FactoMineR is nice because it's easy to pull out the result objects
pca

pca$eig

fviz_eig(pca, ncp = 40)
```
# Label categories

```{r}
clust <- tibble(cluster = km$cluster) %>%
  mutate(category = case_when(cluster == 1 ~ "Modern Wargame",
                              cluster == 2 ~ "Misc",
                              cluster == 3 ~ "Children's game",
                              cluster == 4 ~ "Party card game",
                              cluster == 5 ~ "Economic",
                              cluster == 6 ~ "Card game",
                              cluster == 7 ~ "Science fiction",
                              cluster == 8 ~ "Medieval",
                              cluster == 9 ~ "Ancient",
                              cluster == 10 ~ "Abstract puzzle",
                              cluster == 11 ~ "Action game",
                              cluster == 12 ~ "Classical wargame")) %>%
  bind_cols(deet3 %>% select(id:minage)) %>%
  left_join(rat, by = "id")

table(clust$category)

# Check top rated games in each category

clust %>%
  group_by(category) %>%
  arrange(desc(users_rated)) %>%
  slice(1:5)
```

## Explore categories

```{r}
library(ggridges)

theme_set(theme_minimal())

clust %>%
  ggplot(aes(y = fct_reorder(category, average, mean), 
            # really useful function - reorder category based on the mean of average
             x = average, fill = stat(x))) +
    geom_density_ridges_gradient() +
    scale_fill_viridis_c() +
    labs(y = "", x = "Average Rating") +
    theme(legend.position = "none")

clust %>%
  ggplot(aes(y = fct_reorder(category, age, mean),
             x = age, fill = stat(x))) +
    geom_density_ridges_gradient() +
    scale_fill_viridis_c() +
    scale_x_continuous(limits = c(0, 50))

clust %>%
  ggplot(aes(y = fct_reorder(category, minage, mean),
             x = minage, fill = stat(x))) +
    geom_density_ridges_gradient() +
    scale_fill_viridis_c()

clust %>%
  ggplot(aes(y = fct_reorder(category, users_rated, mean),
             x = users_rated, fill = stat(x))) +
    geom_density_ridges_gradient() +
    scale_fill_viridis_c() +
    scale_x_continuous(limits = c(0, 500))
```

```{r}
clust %>%
  filter(age < 100) %>%
  ggplot(aes(x = age, y = average)) +
    geom_jitter(alpha = 0.1, shape = 16) +
    geom_smooth()

clust %>%
  ggplot(aes(x = users_rated, y = average)) +
    geom_jitter(alpha = 0.1, shape = 16)
```

```{r}
library(ggimage)

# set font
font = "Bahnschrift"

high_rank <- clust %>%
  group_by(category) %>%
  mutate(avg_score = mean(average, na.rm = TRUE)) %>%
  arrange(rank) %>%
  slice(1)

high_rank %>%
  ggplot(aes(x = average, y = reorder(category, avg_score))) +
    geom_point(aes(x = avg_score), shape = 15, color = "blue") +
    geom_image(aes(image = thumbnail), size = .1) +
    geom_text(aes(label = primary, x = average + .5), 
              size = 3, hjust = 0, alpha = 0.5) +
    scale_x_continuous(limits = c(5, 11)) +
    labs(title = "Highest ranked board games by category",
         x = "Average Rating", y = "") +
    annotate("text", x = 6.8, y = 12, hjust = 1,
             label = "Category average rating", color = "blue",
             family = font, size = 4) +
    theme(text = element_text(family = font))
```

Overrated games - most user ratings while the average rating is below one standard deviation of the category mean

```{r}
overrate <- clust %>%
  group_by(category) %>%
  mutate(avg_score = mean(average, na.rm = TRUE),
         sd_score = sd(average, na.rm = TRUE)) %>%
  filter(average < avg_score - sd_score) %>%
  arrange(desc(users_rated)) %>%
  slice(1)

# combine data into one chart

g <- overrate %>%
  ggplot(aes(x = average, y = reorder(category, avg_score))) +
    geom_point(aes(x = avg_score), shape = 1, color = "orangered",
               size = 2) +
    geom_image(aes(image = thumbnail), size = .1) +
    geom_text(aes(label = primary, x = average - .6), 
              size = 2.5, hjust = 1, alpha = 0.5) +
    geom_image(data = high_rank, aes(image = thumbnail), size = .1) +
    geom_text(data = high_rank, aes(label = primary, x = average + .6), 
              size = 2.5, hjust = 0, alpha = 0.5) +
    scale_x_continuous(limits = c(2, 12),
                       breaks = c(2, 4, 6, 8, 10)) +
    labs(title = "Highest rated and overrated games by category",
         subtitle = "@rbamattre - Source: BoardGameGeek",
         x = "Average Rating", y = "") +
    #annotate("text", x = 7.2, y = 12, hjust = 0,
    #         label = "Category average rating", color = "blue",
    #         family = font, size = 4) +
    theme(text = element_text(family = font))

g

ggsave("2022-01-25 board games.png", g, width = 8, height = 6, dpi = 150,
       bg = "white")
```
